---
title: "Untitled"
author: "greg ellison"
date: "2/2/2022"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

data <- read.csv("data/StormDataWithZScores.csv")
```

```{r cutoffs, include=FALSE}
sscore_cuts <- c(64, 82, 95, 112, 136)

s <- sd(data$MaxWindSpeed,na.rm=TRUE)
m <- mean(data$MaxWindSpeed,na.rm=TRUE)

# z-score categories base on saff-simp (windspeeds)
zscore_cuts <- (sscore_cuts - m) / s

# Surge: z-score category cuts based on surge saff-simp
surge_cuts <- c(4,5,8,12,18)
s <- sd(data$Surge,na.rm=TRUE)
m <- mean(data$Surge,na.rm = TRUE)
s_zscore_cuts <- (surge_cuts - m) / s

# Pressure: z-score category cuts based on pressure saff-simp
pressure_cuts <- -1 * c(990,979,965,945,920)
s <- sd(-1 * data$Pressure,na.rm=TRUE)
m <- mean(-1 * data$Pressure,na.rm = TRUE)
p_zscore_cuts <- (pressure_cuts - m) / s

# z-score cutoffs by averaging wind,pressure,surge cutoffs:
avgcuts <- (1/3) * (p_zscore_cuts + s_zscore_cuts + zscore_cuts)


data$WindCat <-     .bincode(data$MaxWindSpeed_Z, c(-5,zscore_cuts,10))-1
data$RainCat <-     .bincode(data$RainfallPointMax_Z, c(-5,zscore_cuts,10))-1
data$SurgeCat <-    .bincode(data$Surge_Z, c(-5,zscore_cuts,10))-1
data$PressureCat <- .bincode(-1 * data$Pressure_Z, c(-5,zscore_cuts,10))-1
data$TornadoCat <-  .bincode(data$Tornado_Z, c(-5,zscore_cuts,10))-1
data$RadiusCat <- .bincode(data$Radius_Z, c(-5,zscore_cuts,10))-1

data$WindCatSurgeCuts     <- .bincode(data$MaxWindSpeed_Z, c(-5,s_zscore_cuts,10))-1
data$RainCatSurgeCuts     <- .bincode(data$RainfallPointMax_Z, c(-5,s_zscore_cuts,10))-1
data$SurgeCatSurgeCuts    <- .bincode(data$Surge_Z, c(-5,s_zscore_cuts,10))-1
data$PressureCatSurgeCuts <- .bincode(-1 * data$Pressure_Z, c(-5,s_zscore_cuts,10))-1
data$TornadoCatSurgeCuts  <- .bincode(data$Tornado_Z, c(-5,s_zscore_cuts,10))-1
data$RadiusCatSurgeCuts   <- .bincode(data$Radius_Z, c(-5,s_zscore_cuts,10))-1

data$WindCatPressureCuts     <- .bincode(data$MaxWindSpeed_Z, c(-5,p_zscore_cuts,10))-1
data$RainCatPressureCuts     <- .bincode(data$RainfallPointMax_Z, c(-5,p_zscore_cuts,10))-1
data$SurgeCatPressureCuts    <- .bincode(data$Surge_Z, c(-5,p_zscore_cuts,10))-1
data$PressureCatPressureCuts <- .bincode(-1 * data$Pressure_Z, c(-5,p_zscore_cuts,10))-1
data$TornadoCatPressureCuts  <- .bincode(data$Tornado_Z, c(-5,p_zscore_cuts,10))-1
data$RadiusCatPressureCuts   <- .bincode(data$Radius_Z, c(-5,p_zscore_cuts,10))-1

data$RainCatAvgCuts     <- .bincode(data$RainfallPointMax_Z, c(-5,avgcuts,10))-1
data$PressureCatAvgCuts <- .bincode(data$Pressure_Z, c(-5,avgcuts,10))-1
data$TornadoCatAvgCuts  <- .bincode(data$Tornado_Z, c(-5,avgcuts,10))-1
data$RadiusCatAvgCuts   <- .bincode(data$Radius_Z, c(-5,avgcuts,10))-1
```

# Z-Score Cutoffs

1. Original Cuts -- Z-Scores for Windspeed Saffir-Simpson Scale
2. Pressure Cuts -- Z-Scores for Pressure "Saffir-Simpson" Scale
3. Surge Cuts -- Z-Scores for Surge "Saffir-Simpson" Scale
4. Averaged Cuts -- Avg. of the three above Z-Score cuts

(Reference: https://www.nps.gov/articles/saffir-simpson-hurricane-scale.htm)

```{r cut_table}
tab <- rbind(zscore_cuts,s_zscore_cuts,p_zscore_cuts,avgcuts)
row.names(tab) <- c("Windspeed","Surge","Pressure","Average")
names(tab) <- c("Cat 1", "Cat 2", "Cat 3", "Cat 4", "Cat 5")
kable(tab, col.names = c("Cat 1", "Cat 2", "Cat 3", "Cat 4", "Cat 5"))
```

## Windspeed Categories
### Based on Windspeed Saffir-Simpson Z-scores
```{r wind_hist}
hist(data$MaxWindSpeed_Z,breaks=seq(-2.5,3.5,length.out=15))
abline(v=zscore_cuts,lwd=3,col="blue")
tab <- table(data$WindCat)
tab <- cbind(tab, tab/sum(tab))
kable(tab, col.names = c("Count", "Rate"))
```

## Pressure Categories
### Based on Pressure "Saffir-Simpson" z-scores 
```{r pressure_hist}
hist(data$Pressure_Z,breaks=15)
abline(v=p_zscore_cuts,lwd=3,col="blue")
abline(v=zscore_cuts,col="gray50")
tab <- table(data$PressureCatPressureCuts)
tab <- cbind(tab, tab/sum(tab))
kable(tab, col.names = c("Count", "Rate"))
```

## Surge Categories
### Based on Surge "Saffir-Simpson" z-scores 
```{r surge_hist}
hist(data$Surge_Z,breaks=15)
abline(v=s_zscore_cuts,lwd=3,col="blue")
abline(v=zscore_cuts,col="gray50")
tab <- table(data$SurgeCatSurgeCuts)
tab <- cbind(tab, tab/sum(tab))
kable(tab, col.names = c("Count", "Rate"))
```


## Rainfall Categories
### Based on Windspeed Saffir-Simpson and Averaged z-score cutoffs
```{r rain_hist}
hist(data$RainfallPointMax_Z,breaks=15)
abline(v=avgcuts,lwd=3,col="blue")
abline(v=zscore_cuts,col="gray50")
tab <- table(data$RainCatAvgCuts)
tab <- cbind(tab, tab/sum(tab))
tab2 <- table(data$RainCat)
tab2 <- cbind(tab2, tab2/sum(tab2))
kable(cbind(tab,tab2), col.names = c("Count (Avg.Cuts)", "Rate (Avg.Cuts)", "Count (Orig.Cuts)", "Rate (Orig.Cuts"))
```





## Tornado Categories
### Based on Windspeed Saffir-Simpson and Averaged z-score cutoffs

```{r tornado_hist}
hist(data$Tornado_Z,breaks=15)
abline(v=avgcuts,lwd=3,col="blue")
abline(v=zscore_cuts,col="gray50")
tab <- table(data$TornadoCatAvgCuts)
tab <- cbind(tab, tab/sum(tab))
tab2 <- table(data$TornadoCat)
tab2 <- cbind(tab2, tab2/sum(tab2))
kable(cbind(tab,tab2), col.names = c("Count (Avg.Cuts)", "Rate (Avg.Cuts)", "Count(Orig.Cuts)", "Rate (Orig.Cuts"))
```


## Rainfall Categories
### Based on Windspeed Saffir-Simpson and Averaged z-score cutoffs
```{r radius_hist}
hist(data$Radius_Z,breaks=15)
abline(v=avgcuts,lwd=3,col="blue")
abline(v=zscore_cuts,col="gray50")
tab <- table(data$RadiusCatAvgCuts)
tab <- cbind(tab, tab/sum(tab))
tab2 <- table(data$RadiusCat)
tab2 <- cbind(tab2, tab2/sum(tab2))
kable(cbind(tab,tab2), col.names = c("Count (Avg.Cuts)", "Rate (Avg.Cuts)", "Count(Orig.Cuts)", "Rate (Orig.Cuts"))
```


```{r include=FALSE}
hist(data$Surge_Z)
abline(v = zscore_cuts)

hist(data$Tornado_Z)
abline(v = zscore_cuts)

hist(data$Pressure)
hist(data$Pressure_Z)
abline(v = zscore_cuts)

# Surge Changes:
table(data$SurgeCat)
table(data$SurgeCatSurgeCuts)
table(data$SurgeCat,data$SurgeCatSurgeCuts)


# Pressure Changes
table(data$PressureCat)
table(data$PressureCatPressureCuts)
table(data$PressureCat,data$PressureCatPressureCuts)


# Rainfall?
# https://www.ncei.noaa.gov/news/scale-hurricane-rainfall-magnitude

data$RainCatAvgCuts     <- .bincode(data$RainfallPointMax_Z, c(-5,avgcuts,10))-1

table(data$RainCat)
table(data$RainCatSurgeCuts)
table(data$RainCatPressureCuts)
table(data$RainCatAvgCuts)

# https://www.nps.gov/articles/saffir-simpson-hurricane-scale.htm#:~:text=In%20the%20early%201970s%2C%20Herbert,intensity%20from%201%20to%205.&text=Low%20atmospheric%20pressure%20is%20also%20associated%20with%20hurricanes.
```


```{r}
source("R/plot_functions.R")

cuts <- list(ws_cuts    = c(-10,zscore_cuts,10), 
             surge_cuts = c(-10,s_zscore_cuts,10), 
             pres_cuts  = c(-10,p_zscore_cuts,10), 
             avg_cuts   = c(-10,avgcuts,10))

h <- data[15,]
make_bar_plot(h, cuts) 
```


```{r}


data <- data %>% 
  mutate(RainCatAvgCuts = ifelse(is.na(RainCatAvgCuts), 0, RainCatAvgCuts),
         RainCat = ifelse(is.na(RainCat), 0, RainCat),
         RadiusCatAvgCuts = ifelse(is.na(RadiusCatAvgCuts), 0, RadiusCatAvgCuts),
         RadiusCat = ifelse(is.na(RadiusCat), 0, RadiusCat),
         TornadoCatAvgCuts = ifelse(is.na(TornadoCatAvgCuts), 0, TornadoCatAvgCuts),
         TornadoCat = ifelse(is.na(TornadoCat), 0, TornadoCat),
         SurgeCatSurgeCuts = ifelse(is.na(SurgeCatSurgeCuts), 0, SurgeCatSurgeCuts),
         SurgeCat = ifelse(is.na(SurgeCat), 0, SurgeCat),
         PressureCat = ifelse(is.na(PressureCat), 0, PressureCat),
         PressureCatPressureCuts = ifelse(is.na(PressureCatPressureCuts), 0, PressureCatPressureCuts))  %>% 
  mutate(RainCatDiff = RainCatAvgCuts != RainCat,
         SizeCatDiff = RadiusCatAvgCuts != RadiusCat,
         TornadoCatDiff = TornadoCatAvgCuts != TornadoCat,
         SurgeCatDiff = SurgeCatSurgeCuts != SurgeCat,
         PressureCatDiff = PressureCatPressureCuts != PressureCat,
         NDiff = RainCatDiff + SizeCatDiff + TornadoCatDiff + SurgeCatDiff + PressureCatDiff,
         TotalCat = RainCat + RadiusCat + TornadoCat + SurgeCat + PressureCat
         )

data_diffs <- data %>% 
  filter(NDiff >= 1 & TotalCat >= 9)
data_diffs
```
